Live Functional Programming with Typed Holes

### 2018-11-20 hazelnut-dynamics.tex
started making ArXiV version and updated canonical indet forms statement
add more specific references to sections of the appendix behind ifarxiv flag
added country to author affiliations per ACM request + fixed a spelling mistake in related work (both submitted in final version)

\PassOptionsToPackage{svgnames,dvipsnames,svgnames}{xcolor}
\newif\ifarxiv
\arxivtrue
\ifarxiv
\documentclass[acmsmall,screen,nonacm]{acmart}
\else
%%% Note: arxiv does not want line numbers (they are detected somehow, and are not allowed)
\documentclass[acmsmall,screen]{acmart}
% \settopmatter{printfolios=false,printccs=false,printacmref=false}
% \settopmatter{printccs=false,printacmref=false}
\fi

@@

%% Copyright information
%% Supplied to authors (based on authors' rights management selection;
%% see authors.acm.org) by publisher for camera-ready submission
% \setcopyright{none}             %% For review submission
%\setcopyright{acmcopyright}
%\setcopyright{acmlicensed}
%\setcopyright{rightsretained}
%\copyrightyear{2017}           %% If different from \acmYear
\setcopyright{rightsretained} 
\acmJournal{PACMPL}
\acmYear{2019} \acmVolume{3} \acmNumber{POPL} \ifarxiv \acmArticle{1} \else \acmArticle{14} \fi \acmMonth{1} \acmPrice{}\acmDOI{10.1145/3290327}
\copyrightyear{2019}

@@

%% Title information
\title{Live Functional Programming with Typed Holes}         %% [Short Title] is optional;
\subtitle{Extended Version}
\subtitlenote{The original version of this article appears as ...}
                                        %% when present, will be used in
                                        %% header instead of Full Title.
% \titlenote{with title note}             %% \titlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'
% \subtitle{Subtitle}                     %% \subtitle is optional
% \subtitlenote{with subtitle note}       %% \subtitlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'

@@

%% Author with single affiliation.
\author{Cyrus Omar}
% \authornote{with author1 note}          %% \authornote is optional;
                                        %% can be repeated if necessary
% \orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
\affiliation{
  % \position{Position1}
  % \department{Department1}              %% \department is recommended
  \institution{University of Chicago}            %% \institution is required
  % \streetaddress{Street1 Address1}
  % \city{City1}
  % \state{State1}
  % \postcode{Post-Code1}
  % \country{Country1}
}
\email{comar@cs.uchicago.edu}          %% \email is recommended

\author{Ian Voysey}
% \authornote{with author1 note}          %% \authornote is optional;
                                        %% can be repeated if necessary
% \orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
\affiliation{
  % \position{Position1}
  % \department{Department1}              %% \department is recommended
  \institution{Carnegie Mellon University}            %% \institution is required
  % \streetaddress{Street1 Address1}
  % \city{City1}
  % \state{State1}
  % \postcode{Post-Code1}
  % \country{Country1}
}
\email{iev@cs.cmu.edu}          %% \email is recommended

\author{Ravi Chugh}
% \authornote{with author1 note}          %% \authornote is optional;
                                        %% can be repeated if necessary
% \orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
\affiliation{
  % \position{Position1}
  % \department{Department1}              %% \department is recommended
  \institution{University of Chicago}            %% \institution is required
  % \streetaddress{Street1 Address1}
  % \city{City1}
  % \state{State1}
  % \postcode{Post-Code1}
  % \country{Country1}
}
\email{rchugh@cs.uchicago.edu}          %% \email is recommended

\author{Matthew A. Hammer}
% \authornote{with author1 note}          %% \authornote is optional;
                                        %% can be repeated if necessary
% \orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
\affiliation{
  % \position{Position1}
  % \department{Department1}              %% \department is recommended
  \institution{University of Colorado Boulder}            %% \institution is required
  % \streetaddress{Street1 Address1}
  % \city{City1}
  % \state{State1}
  % \postcode{Post-Code1}
  % \country{Country1}
}
\email{matthew.hammer@colorado.edu}          %% \email is recommended

@@

%\clearpage
\bibliography{references,all.short,hazel_NSF}

>>>>

\PassOptionsToPackage{svgnames,dvipsnames,svgnames}{xcolor}
\newif\ifarxiv
\arxivtrue
% \arxivfalse
\ifarxiv
\documentclass[acmsmall,screen,nonacm]{acmart}
\else
%%% Note: arxiv does not want line numbers (they are detected somehow, and are not allowed)
\documentclass[acmsmall,screen]{acmart}
% \settopmatter{printfolios=false,printccs=false,printacmref=false}
% \settopmatter{printccs=false,printacmref=false}
\fi

@@

\ifarxiv
\setcopyright{rightsretained} 
\acmJournal{PACMPL}
\acmYear{2019} \acmVolume{3} \acmNumber{POPL} \ifarxiv \acmArticle{1} \else \acmArticle{14} \fi \acmMonth{1} \acmPrice{}\acmDOI{10.1145/3290327}
\copyrightyear{2019}
\else
%%% The following is specific to POPL '19 and the paper
%%% 'Live Functional Programming with Typed Holes'
%%% by Cyrus Omar, Ian Voysey, Ravi Chugh, and Matthew A. Hammer.
%%%
\setcopyright{rightsretained}
\acmPrice{}
\acmDOI{10.1145/3290327}
\acmYear{2019}
\copyrightyear{2019}
\acmJournal{PACMPL}
\acmVolume{3}
\acmNumber{POPL}
\acmArticle{14}
\acmMonth{1}
\fi

@@

%% Title information
\title{Live Functional Programming with Typed Holes}         %% [Short Title] is optional;
\ifarxiv
\subtitle{Extended Version}
\subtitlenote{The original version of this article was published in the POPL 2019 edition of PACMPL \cite{HazelnutLive}. This extended version includes an additional appendix.}
\fi
                                        %% when present, will be used in

                                        %% header instead of Full Title.
% \titlenote{with title note}             %% \titlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'
% \subtitle{Subtitle}                     %% \subtitle is optional
% \subtitlenote{with subtitle note}       %% \subtitlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'

@@

%% Author with single affiliation.
\author{Cyrus Omar}
% \authornote{with author1 note}          %% \authornote is optional;
                                        %% can be repeated if necessary
% \orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
\affiliation{
  % \position{Position1}
  % \department{Department1}              %% \department is recommended
  \institution{University of Chicago, USA}            %% \institution is required
  % \streetaddress{Street1 Address1}
  % \city{City1}
  % \state{State1}
  % \postcode{Post-Code1}
  % \country{Country1}
}
\email{comar@cs.uchicago.edu}          %% \email is recommended

\author{Ian Voysey}
% \authornote{with author1 note}          %% \authornote is optional;
                                        %% can be repeated if necessary
% \orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
\affiliation{
  % \position{Position1}
  % \department{Department1}              %% \department is recommended
  \institution{Carnegie Mellon University, USA}            %% \institution is required
  % \streetaddress{Street1 Address1}
  % \city{City1}
  % \state{State1}
  % \postcode{Post-Code1}
  % \country{Country1}
}
\email{iev@cs.cmu.edu}          %% \email is recommended

\author{Ravi Chugh}
% \authornote{with author1 note}          %% \authornote is optional;
                                        %% can be repeated if necessary
% \orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
\affiliation{
  % \position{Position1}
  % \department{Department1}              %% \department is recommended
  \institution{University of Chicago, USA}            %% \institution is required
  % \streetaddress{Street1 Address1}
  % \city{City1}
  % \state{State1}
  % \postcode{Post-Code1}
  % \country{Country1}
}
\email{rchugh@cs.uchicago.edu}          %% \email is recommended

\author{Matthew A. Hammer}
% \authornote{with author1 note}          %% \authornote is optional;
                                        %% can be repeated if necessary
% \orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
\affiliation{
  % \position{Position1}
  % \department{Department1}              %% \department is recommended
  \institution{University of Colorado Boulder, USA}            %% \institution is required
  % \streetaddress{Street1 Address1}
  % \city{City1}
  % \state{State1}
  % \postcode{Post-Code1}
  % \country{Country1}
}
\email{matthew.hammer@colorado.edu}          %% \email is recommended

@@

%\clearpage
\bibliography{all.short}

### 2018-11-20 related-work.tex
added country to author affiliations per ACM request + fixed a spelling mistake in related work (both submitted in final version)

We use the machinery 
borrowed from CMTT only extralinguistically.
%
A key feature of CMTT that we have not yet touched on is the
\emph{internalization} of metavariable binding and contextual
substitution via the contextual modal types, $[\hGamma]\htau$, which
are introduced by the operation $\mathsf{box}(\hGamma.d)$ and
eliminated by the operation $\mathsf{letbox}(d_1, u.d_2)$.
%
A program with expression holes can be interpretted as being bound under a number of these $\mathsf{letbox}$ constructs, i.e. CMTT serves as a ``{development calculus}'' \cite{DBLP:phd/ethos/McBride00}. Live programming corresponds to reduction under the metavariable binders interleaved with elimination steps. 
%
This suggests the possibility of \emph{computing} hole fillings by specifying a non-trivial expression of the appropriate contextual 
modal type in this development calculus.
%
This could, in turn, serve as the basis for a computational hole
refinement system that supports efficient live programming, extending the capabilities of purely static hole
refinement systems like those available in many languages,
e.g. the editor-integrated system of Idris
\cite{brady2013idris,Korkut:2018:ETE:3240719.3241791} and the hole refinement system of Beluga
\cite{DBLP:conf/flops/Pientka10,pientka2015inductive}.
%
Each applied hole filling can be interpretted as inducing a new dynamic
\emph{edit stage}.
%
This contextual modal interpretation of live hole refinement
nicely mirrors the modal interpretation of staging and partial evaluation
\cite{Davies:2001op}, with indeterminate forms corresponding to the residual programs that arise when performing partial evaluation. 
%
The difference is that staging and partial evaluation systems evaluate around an input that sits outside of a
function \cite{Jones:1993uq}, whereas holes are contextual, i.e. they are located inside the program.

>>>>

We use the machinery 
borrowed from CMTT only extralinguistically.
%
A key feature of CMTT that we have not yet touched on is the
\emph{internalization} of metavariable binding and contextual
substitution via the contextual modal types, $[\hGamma]\htau$, which
are introduced by the operation $\mathsf{box}(\hGamma.d)$ and
eliminated by the operation $\mathsf{letbox}(d_1, u.d_2)$.
%
A program with expression holes can be interpreted as being bound under a number of these $\mathsf{letbox}$ constructs, i.e. CMTT serves as a ``{development calculus}'' \cite{DBLP:phd/ethos/McBride00}. Live programming corresponds to reduction under the metavariable binders interleaved with elimination steps. 
%
This suggests the possibility of \emph{computing} hole fillings by specifying a non-trivial expression of the appropriate contextual 
modal type in this development calculus.
%
This could, in turn, serve as the basis for a computational hole
refinement system that supports efficient live programming, extending the capabilities of purely static hole
refinement systems like those available in many languages,
e.g. the editor-integrated system of Idris
\cite{brady2013idris,Korkut:2018:ETE:3240719.3241791} and the hole refinement system of Beluga
\cite{DBLP:conf/flops/Pientka10,pientka2015inductive}.
%
Each applied hole filling can be interpreted as inducing a new dynamic
\emph{edit stage}.
%
This contextual modal interpretation of live hole refinement
nicely mirrors the modal interpretation of staging and partial evaluation
\cite{Davies:2001op}, with indeterminate forms corresponding to the residual programs that arise when performing partial evaluation. 
%
The difference is that staging and partial evaluation systems evaluate around an input that sits outside of a
function \cite{Jones:1993uq}, whereas holes are contextual, i.e. they are located inside the program.
